---
title: "Welcome! Expect Software, Stats, Hoops and Trash."
description: |
  Statistics, sports and recent gibberish are all moving targets. Hence I’ll do my best to dish out a monthly editorial like update. Plus, doing so is an interesting mechanism to check whether I am sticking to plans made a month ago.
author:
  - name: Matt Bannert
    url: https://whatsgood.io
    affiliation: ETH Zurich
    affiliation_url: https://ethz.ch
date: 10-26-2021
output:
  distill::distill_article:
    self_contained: false
twitter:
  site: "@rstudio"
  creator: "@whatsgoodio"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating blogs with Distill at:
# https://rstudio.github.io/distill/blog.html

```

The end of 2021 has been busy for me. I have been working on a 3-part mini series of the [RSE stories]() podcast. RSE Stories is the podcast of the [U.S. society of research software engineering](). I am looking forward to interesting discussions but I am also looking forward to the hosting and audio editing experience.

In fall, my PhD level course went into its second year. My teaching at ETH is part of my `r emojifont::emoji("heart")`-project [Hacking for Social Sciences](https://h4sci.github.io/) in which I teach data automation for researchers. Looking back, one of the most interesting things
about this project is that I meant to call it *Hacking for Economists*. And if it wasn't for @ (thank you, sir!)  that would have been the name and many very appreciated contributors might have thought *not-for-me* before they even started. The wide spread interest and feedback from across fields has fueled the project and will eventually change its name to something more inclusive like  *Hacking for Researchers*. As you might have seen on twitter, I have teamed up with @crcgrubbsd and @CRC_MathStats to come up with an actual book. Here's a work-in-progress version of my book https://h4sci.github.io/h4sci-book/ . (Feedback welcome!)

```{r}
library(tweetrmd)
tweet_embed("https://twitter.com/whatsgoodio/status/1460525798369206272")

```

And, oh boi, did the hoop season start. Dark horse teams, surprises and several seasons worth of interesting is-team-X-for-real stories to follow after about 20 games into the NBA season. The Bulls are back, the champion Bucks manage to crawl out of the hole injuries dug for them, Miami's OG front office put together an off-season of its own class and the West is wild as it ever was. The Warriors are back looking high-octane (and Klay is not even back), the Suns show that they are not a fluke, the Jazz seem to be in tune will the rest of the West is still busy sorting their feet. Also on an individual level, there is a lot of development worth following. The rise of Ja Morant continues while players struggle under the rule changes (or is it the new ball?). Hey, after all these years of Spalding, it's even irritating to spectators. 

With the year coming to an end and the holiday season ahead (including darts and more basketball), I plan to tone the programming stuff down a little and explore new projects. That is more podcasting ahead, more book writing and looking into an online marketing project with @jburkhardt. 

Have a peaceful holiday season

Matt


