---
title: "Talks"
description: |
  I enjoy exchange with students, partners from the public and private sector as well as with the open source community at large. On this site I keep track of my blabla  and related material such as slides or videos.
---


### The {kofdata} API wrapper R package (adminR Lugano, 2021)

Presentation of the {kofdata} R package at the adminR meetup during Statistiktage 2021. {kofdata} is an R layer around the KOF data REST API for data transder. 

[slides](https://matt-talks.github.io/adminR-lugano-2021/#2) | [video](https://www.youtube.com/watch?v=VJC-VU4g1KA) 


### The {timeseriesdb} R package (uRos, 2020)

[slides](https://matt-talks.github.io/timeseriesdb-uros/#1) | [video](https://www.youtube.com/watch?v=8sAZE8fYvNU) 


### From Airflow to Docker (uRos, 2020)

Lightning presentation held at uRos - R in Official statistics. I talk about R in an Open Source Production Environment.

[slides](https://matt-talks.github.io/uros2020/#1) | [video](https://youtu.be/bTYrw1RA45Y?t=2449)


### R Alonside Airflow, Docker and Gitlab CI (eRum, 2020)

I presented the Open Source ecosytem we have set up at KOF Swiss Economic Institute and discussed how R behaves in this environment. The talk was given as part of the in production track at eRum 2020. 

[video](https://youtu.be/VQtEI-akVyk)


### useR! 2019 Travel Report (ZH R Users Meetup, 2019)

The #train2useR journey started at Zurich main station. Together with other R enthusiasts, I took the train to Toulouse, France. When we were back in Zurich we had plenty to report from decent donuts, to superb talks and wrecked trains. 


```{r,echo=FALSE}
library(tweetrmd)
tweetrmd::include_tweet("https://twitter.com/whatsgoodio/status/1148310157325152257")
```

[event](https://www.meetup.com/Zurich-R-User-Group/events/265175210/) | [slides](https://zhr-meetup-201910.netlify.app/#1)


### useR! 2019: The {timeseriesdb} R package

At useR! 2019 in Toulouse, France, I presented plans for a new release of the {timeseriesdb} R package. Fun fact, I was the first to come up with a user2019 subdomain at netlify ;) 

[slides](https://user2019.netlify.app/#1) | [video](https://www.youtube.com/watch?v=4Y-7B9aIP6Y)


### techTalk  (invited, Aarau, 2019)

I discussed *R in Production - The R Ecosystem @KOF* 
with an audience from cantonal administration of the Swiss Canton of Aargau. 


### adminR: Economic Statistics (Zurich, 2019)

```{r, echo=FALSE}
tweetrmd::include_tweet("https://twitter.com/adminR3/status/1106334380170113024")
```

I co-hosted adminR 2019 and gave the host talk at adminR - Economic Statistics. 

[event](https://www.meetup.com/adminR/)

### Schweizer Statistiktage: Open Source Driven Data Pipelines (zurich, 2018)



