---
title: "It's on CRAN - Introducing the kofdata R package"
author: admin
date: 2018-12-01T21:13:14-05:00
categories: ["R","API","PostgreSQL","Official Statistics"]
tags: ["R Markdown", "plot", "regression"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

At KOF, we have recently introduced a REST Interface to conveniently automate 
data downloads. R users can now benefit from an even more convenient way to consume KOF data:
the brand new 'kofdata' CRAN package. The package allows to download Swiss
time series data and meta information through simple R functions. 




Installing the package is a matter of:


```{r, eval=FALSE}
install.packages("kofdata")
```

Once you have installed the package you can download data and obtain R time 
series objects with a simple function call. Get Switzerland's most popular macro
economic *forward looking indicator*, the KOF Barometer like this: 


```{r, eval=FALSE}
library(kofdata)
get_time_series("kofbarometer")
```

Because you will receive an R time series object you can easily create a 
time series plot out-of-the-box using base R's plot. Or get our 
[tstools](https://github.com/mbannert/tstools) package and given even neater plots... 


```{r, eval=TRUE, echo=FALSE}
library(kofdata)
```


```{r}
# base R alternative without additional plot packages
tslist <- get_time_series("kofbarometer")
plot(tslist$kofbarometer)

library(tstools)
tsplot(tslist)
```

And yes, there's a bit more to the package than just querying single time series. 
Here is a small overview of the current, basic functionality.

```{r}
lsf.str("package:kofdata")
```











